---
title: "Power tables for t-test"
output: html_document
date: "2025-06-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(rmutil)
library(kableExtra)
library(extraDistr)
set.seed(1)
rm(list=ls())
```

# Power table

We evaluate the ability of the t-test to reject the null hypothesis when it is false in a series of experiments where the population distribution is non-normal and the sample size varies. Each cell of the table represents the proportion of rejections out of 10,000 samples taken.

Let
\[ H_0: \mu = 5\]
\[ H_a: \mu \ge 5  \]

and ($\mu = 7$) for each of the below distributions. 

```{r}
n <- c(seq(10,100, by=10))

dist_funs <- list(
  Normal    = function(n) rnorm(n, mean = 7, sd = 1),
  t_shifted_7= function(n) rt(n, df = 5) + 7,  # centered at 7
  Cauchy    = function(n) rcauchy(n, location = 7, scale = 1),
  LogNormal = function(n) rlnorm(n, meanlog = log(7), sdlog = 0.25), # approx center near 7
  Exponential = function(n) rexp(n, rate = 1/7),  # mean = 7
  Uniform = function(n) runif(n, min = 0, max=14),
  Laplace   = function(n) rmutil::rlaplace(n, m = 7, s = 1),
  ChiSq      = function(n) rchisq(n, df = 4) + (7 - 4),                # right-skewed, mean ≈ 7
  Triangular = function(n) extraDistr::rtriang(n, a = 0, b = 14, c = 7), # symmetric, bounded
  Bimodal    = function(n) sample(c(rnorm(n/2, 5, 1), rnorm(n/2, 9, 1))) # symmetric bimodal
)

power_table <- as.data.frame(matrix(NA, nrow = length(dist_funs), ncol = length(n)))
colnames(power_table) <- n
rownames(power_table) <- names(dist_funs)
```
```{r}
for(distrib in names(dist_funs)){
  call <- dist_funs[[distrib]]
  
  for(ss in n){
    # cat(distrib, ss, "\n")
    samples <- replicate(100, call(ss))
    p_vals <- apply(samples, 2, function(x) {t.test(x, alternative = "greater", mu = 5)$p.value})
    prop_reject <- mean(p_vals < .05)
    power_table[distrib, which(n==ss)] <- prop_reject
  }
}

kable(power_table, digits=2) %>% kable_styling()
```

# Type I error table


We evaluate the tendency of the t-test to reject the null hypothesis when it is true. 

Let
\[ H_0: \mu = 7\]
\[ H_a: \mu \ge 7  \]

and ($\mu = 7$) for each of the below distributions. 

```{r}
type1_table <- as.data.frame(matrix(NA, nrow = length(dist_funs), ncol = length(n)))
colnames(type1_table) <- n
rownames(type1_table) <- names(dist_funs)

for(distrib in names(dist_funs)){
  call <- dist_funs[[distrib]]
  
  for(ss in n){
    # cat(distrib, ss, "\n")
    samples <- replicate(100, call(ss))
    p_vals <- apply(samples, 2, function(x) {t.test(x, alternative = "greater", mu = 7)$p.value})
    prop_reject <- mean(p_vals < .05)
    type1_table[distrib, which(n==ss)] <- prop_reject
  }
}

kable(type1_table, digits=2) %>% kable_styling()
```

